---
title: "Project_DataAnalysis"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Working Directory and Load Data 
```{r, include=FALSE}
# Set your working directory

setwd("/Users/rmlandman/Desktop/Data Analytics/Environmental_Data_Analytics_2020/Project/EDA_2020_Project")
getwd()

# Load your packages


library(plyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(viridis)
library(RColorBrewer)
library(cowplot)

# Set your ggplot theme

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)

# Load your datasets

EC_Flow.Nutrients_Wide <-
  read.csv("./Data/Processed/EC_Flow.Nutrients_Wide_Processed.csv")

Club.Gorman_Flow.Nutrients_Wide <-
  read.csv("./Data/Processed/ClubGorman_Flow.Nutrients_Wide_Processed.csv")

Club.Gorman_Flow_Long <- read.csv("./Data/Processed/ClubGorman_USGSDischarge_Long.csv")

Club.Gorman_Flow_Wide <- read.csv("./Data/Processed/ClubGorman_USGSDischarge_Wide.csv")

Club.Gorman_Flow.Nutrients_Long <-
  read.csv("./Data/Processed/ClubGorman_Flow.Nutrients_Long_Processed.csv")
```

# Format Dates
```{r, include=FALSE}
Club.Gorman_Flow.Nutrients_Wide$Date <- 
  as.Date(Club.Gorman_Flow.Nutrients_Wide$Date, format = "%Y-%m-%d")
class(Club.Gorman_Flow.Nutrients_Wide$Date)

Club.Gorman_Flow_Wide$Date <- 
  as.Date(Club.Gorman_Flow_Wide$Date, format = "%Y-%m-%d")
class(Club.Gorman_Flow_Wide$Date)

Club.Gorman_Flow_Long$Date <- 
  as.Date(Club.Gorman_Flow_Long$Date, format = "%Y-%m-%d")
class(Club.Gorman_Flow_Long$Date)

EC_Flow.Nutrients_Wide$Date <- 
  as.Date(EC_Flow.Nutrients_Wide$Date, format = "%Y-%m-%d")
class(EC_Flow.Nutrients_Wide$Date)

Club.Gorman_Flow.Nutrients_Long$Date <-
  as.Date(Club.Gorman_Flow.Nutrients_Long$Date, format = "%Y-%m-%d")
class(Club.Gorman_Flow.Nutrients_Long$Date)


```

# Format Site Number to Character
```{r, include=FALSE}

Club.Gorman_Flow.Nutrients_Wide$site_no <- 
  as.character(Club.Gorman_Flow.Nutrients_Wide$site_no)
class(Club.Gorman_Flow.Nutrients_Wide$site_no)

Club.Gorman_Flow_Wide$site_no.x <- 
  as.character(Club.Gorman_Flow_Wide$site_no.x)
class(Club.Gorman_Flow_Wide$site_no.x)

Club.Gorman_Flow_Wide$site_no.y <- 
  as.character(Club.Gorman_Flow_Wide$site_no.y)
class(Club.Gorman_Flow_Wide$site_no.y)

Club.Gorman_Flow_Long$site_no <- 
  as.character(Club.Gorman_Flow_Long$site_no, format)
class(Club.Gorman_Flow_Long$site_no)

EC_Flow.Nutrients_Wide$site_no <- 
  as.character(EC_Flow.Nutrients_Wide$site_no)
class(EC_Flow.Nutrients_Wide$site_no)

EC_Flow.Nutrients_Wide$site_no.x <- 
  as.character(EC_Flow.Nutrients_Wide$site_no.x)
class(EC_Flow.Nutrients_Wide$site_no.x)

EC_Flow.Nutrients_Wide$site_no.y <- 
  as.character(EC_Flow.Nutrients_Wide$site_no.y)
class(EC_Flow.Nutrients_Wide$site_no.y)

```
# Flow Summaries 
```{r}
summary(EC_Flow.Nutrients_Wide$Discharge.C)
summary(EC_Flow.Nutrients_Wide$Discharge.G)
```


# Ttest 
 
```{r}
#Is there a difference between upstream and downstream nutrient levels? 
### continuous response (N or P)
### categorical explanatory
```

### Nitrogen
```{r}
summary(EC_Flow.Nutrients_Wide$Nitrogen)

# Evaluate assumption of normal distribution 
shapiro.test((EC_Flow.Nutrients_Wide$Nitrogen))

### shapiro wilke test the null hypothesis is that the data are a normal distribution 
### p value <0.05 shows we reject null and data is not normally distributed
######## "not well approximated by a normal distribution"

ggplot(EC_Flow.Nutrients_Wide, aes(x = Nitrogen)) +
  geom_histogram(binwidth = 10) 

### histogram shows data is not very rightly skewed 

qqnorm(EC_Flow.Nutrients_Wide$Nitrogen); qqline(EC_Flow.Nutrients_Wide$Nitrogen)
## some samples are higher than normal and some are lower

N.onesample <- t.test(EC_Flow.Nutrients_Wide$Nitrogen, mu = 10, alternative = "less")
N.onesample

# Null hypothesis is that mean = 10, alternative is mean is less than 10
# p value >0.05 so we can reject the null the mean =10

# plot 

N.plot <- ggplot(EC_Flow.Nutrients_Wide, aes(x = Nitrogen)) +
  #geom_density(stat = "count", fill = "gray") + #<-shows count so it's more jagged
  geom_density(fill = "gray") +
  geom_vline(xintercept = 10, color = "#238b45", lty = 2, size = 0.9) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
print(N.plot)


```
##### Nitrogen Results 
> Nitrogen measurments in ellerbe creek were not significantly greater than 10 mg/L, the Maximum Contaminent Level for Nitrogen in drinking water (one sample t-test; t= 3.515, df = 139, p < 0.001) 

### Phosphorus
```{r}
summary(EC_Flow.Nutrients_Wide$Phosphorus)

# Evaluate assumption of normal distribution 
shapiro.test((EC_Flow.Nutrients_Wide$Phosphorus))

### shapiro wilke test the null hypothesis is that the data are a normal distribution 
### p value <0.05 shows we reject null and data is not normally distributed
######## "not well approximated by a normal distribution"

ggplot(EC_Flow.Nutrients_Wide, aes(x = Phosphorus)) +
  geom_histogram(binwidth = 2) 

### histogram shows data is not very rightly skewed 

qqnorm(EC_Flow.Nutrients_Wide$Phosphorus); qqline(EC_Flow.Nutrients_Wide$Phosphorus)
## samples are higher than normal distribution

P.onesample <- t.test(EC_Flow.Nutrients_Wide$Phosphorus, mu = 1, alternative = "less")
P.onesample

# Null hypothesis is that mean = 1, alternative is mean is less than 1
# p value <0.05 so we can reject the null the mean = 1

# plot 

P.plot <- ggplot(EC_Flow.Nutrients_Wide, aes(x = Phosphorus)) +
  #geom_density(stat = "count", fill = "gray") + #<-shows count so it's more jagged
  geom_density(fill = "gray") +
  geom_vline(xintercept = 1, color = "#238b45", lty = 2, size = 0.9) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
print(P.plot)


```
##### Phosphorus Results 
> Phosphorus measurments in ellerbe creek were significantly greater than 1 mg/L, the Maximum Contaminent Level for phosphorus in water (one sample t-test; t= 2.5264, df = 242, p = 0.006) 

# Two Sample t-test (Upstream vs. Downstream)
### Nitrogen
```{r}
shapiro.test(Club.Gorman_Flow.Nutrients_Wide$Nitrogen[Club.Gorman_Flow.Nutrients_Wide$Location == "Upstream"])
shapiro.test(Club.Gorman_Flow.Nutrients_Wide$Nitrogen[Club.Gorman_Flow.Nutrients_Wide$Location == "Downstream"])
var.test(Club.Gorman_Flow.Nutrients_Wide$Nitrogen ~ Club.Gorman_Flow.Nutrients_Wide$Location)

### "var.test" test variance 
### asks are the variances eqaul - what is the diff between variances 
### results: variances are significantly different 
### results: violate assumption of normality and equal variance

ggplot(Club.Gorman_Flow.Nutrients_Wide, aes(x = Nitrogen, color = Location)) +
  geom_freqpoly()


N.twosample <- t.test(Club.Gorman_Flow.Nutrients_Wide$Nitrogen ~ Club.Gorman_Flow.Nutrients_Wide$Location)
N.twosample

```

### Phosphorus
```{r}
shapiro.test(Club.Gorman_Flow.Nutrients_Wide$Phosphorus[Club.Gorman_Flow.Nutrients_Wide$Location == "Upstream"])
shapiro.test(Club.Gorman_Flow.Nutrients_Wide$Phosphorus[Club.Gorman_Flow.Nutrients_Wide$Location == "Downstream"])
var.test(Club.Gorman_Flow.Nutrients_Wide$Phosphorus ~ Club.Gorman_Flow.Nutrients_Wide$Location)

### "var.test" test variance 
### asks are the variances eqaul - what is the diff between variances 
### results: variances are significantly different 
### results: violate assumption of normality and equal variance

ggplot(Club.Gorman_Flow.Nutrients_Wide, aes(x = Phosphorus, color = Location)) +
  geom_freqpoly()


P.twosample <- t.test(Club.Gorman_Flow.Nutrients_Wide$Phosphorus ~ Club.Gorman_Flow.Nutrients_Wide$Location)
P.twosample
```

# Wilcox Test (Don't Assume Normality)
### Nitrogen
```{r}
N.onesample.wilcox <- wilcox.test(Club.Gorman_Flow.Nutrients_Wide$Nitrogen, mu = 10)
N.onesample.wilcox
N.twosample.wilcox <- wilcox.test(Club.Gorman_Flow.Nutrients_Wide$Nitrogen ~ Club.Gorman_Flow.Nutrients_Wide$Location)
N.twosample.wilcox
```

### Phosphorus
```{r}
P.onesample.wilcox <- wilcox.test(Club.Gorman_Flow.Nutrients_Wide$Phosphorus, mu = 1)
P.onesample.wilcox
P.twosample.wilcox <- wilcox.test(Club.Gorman_Flow.Nutrients_Wide$Phosphorus ~ Club.Gorman_Flow.Nutrients_Wide$Location)
P.twosample.wilcox
```

# Simple Linear Regression 
>Is flow a significant predictor of nutrient levels? 
>continuous response (N or P)
>continuous predictor (discharge)

###Nitrogen 
```{r}
NitrogenClub.regression <-lm(data = EC_Flow.Nutrients_Wide, Nitrogen ~ Discharge.C)
summary(NitrogenClub.regression)
## adj R squared says 1 % of variance is explained by depth 

cor.test(EC_Flow.Nutrients_Wide$Nitrogen, EC_Flow.Nutrients_Wide$Discharge.C)


NitrogenGorman.regression <-lm(data = EC_Flow.Nutrients_Wide, Nitrogen ~ Discharge.G)
summary(NitrogenGorman.regression)
## adj R squared says 1 % of variance is explained by depth 

cor.test(EC_Flow.Nutrients_Wide$Nitrogen, EC_Flow.Nutrients_Wide$Discharge.G)
```

### Nitrogen log-log 

```{r}
logNitrogenClub.regression <-lm(data = EC_Flow.Nutrients_Wide, log(Nitrogen) ~ log(Discharge.C))
summary(logNitrogenClub.regression)
## EXPLAIN LOG LOG RELATIONSHIP 

logNitrogenGorman.regression <-lm(data = EC_Flow.Nutrients_Wide, log(Nitrogen) ~ log(Discharge.G))
summary(logNitrogenGorman.regression)
## adj R squared says 1 % of variance is explained by depth 

```


###Phosphorus 
```{r}
PhosphorusClub.regression <-lm(data = EC_Flow.Nutrients_Wide, Phosphorus ~ Discharge.C)
summary(PhosphorusClub.regression)
## adj R squared says 8 % of variance is explained by depth 
## significant relationship p < 0.001

cor.test(EC_Flow.Nutrients_Wide$Phosphorus, EC_Flow.Nutrients_Wide$Discharge.C)


PhosphorusGorman.regression <-lm(data = EC_Flow.Nutrients_Wide, Phosphorus ~ Discharge.G)
summary(PhosphorusGorman.regression)
## adj R squared says 9 % of variance is explained by depth
## significant relationship p < 0.001

cor.test(EC_Flow.Nutrients_Wide$Phosphorus, EC_Flow.Nutrients_Wide$Discharge.G)
```

###Phosphorus Log-Log 
```{r}
logPhosphorusClub.regression <-lm(data = EC_Flow.Nutrients_Wide, log(Phosphorus) ~ log(Discharge.C))
summary(logPhosphorusClub.regression)
## adj R squared says 8 % of variance is explained by depth 
## significant relationship p < 0.001


logPhosphorusGorman.regression <-lm(data = EC_Flow.Nutrients_Wide, log(Phosphorus) ~ log(Discharge.G))
summary(logPhosphorusGorman.regression)
## adj R squared says 9 % of variance is explained by depth
## significant relationship p < 0.001

```

# Plot Regression
### Nitrogen
```{r}
Nitrogen_Discharge <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=Discharge.G, y = Nitrogen))+
  geom_point()+
  geom_smooth(method=lm)+
  ylim(0,10)
print(Nitrogen_Discharge)

#log discharge 
Nitrogen_logDischarge <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=log(Discharge.G), y = Nitrogen))+
  geom_point()+
  geom_smooth(method=lm)+
  ylim(0,10)
print(Nitrogen_logDischarge)

#log both 
logNitrogen_logDischarge <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=log(Discharge.G), y = log(Nitrogen)))+
  geom_point()+
  geom_smooth(method=lm)+
  ylim(0,10)
print(logNitrogen_logDischarge)
```

### Phosphorus 
```{r}
Phosphorus_Discharge <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=Discharge.G, y = Phosphorus))+
  geom_point()+
  geom_smooth(method=lm)+
  ylim(0,2.5)
print(Phosphorus_Discharge)

#log discharge 
Phosphorus_logDischarge <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=log(Discharge.G), y = Phosphorus))+
  geom_point()+
  geom_smooth(method=lm)+
  ylim(0,10)
print(Phosphorus_logDischarge)

#log both 
logPhosphorus_logDischarge <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=log(Discharge.G), y = log(Phosphorus)))+
  geom_point()+
  geom_smooth(method=lm)
print(logPhosphorus_logDischarge)
```

# Nutrients over Time 
### Nutrients over time 1980s-2020
```{r}
Nitrogen_Time <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=Date, y = Nitrogen))+
  geom_point()+
  geom_smooth(method=lm)
print(Nitrogen_Time)

 Phosphorus_Time <- 
  ggplot(EC_Flow.Nutrients_Wide, aes (x=Date, y = Phosphorus))+
  geom_point()+
  geom_smooth(method=lm)
print(Phosphorus_Time)
```

### Test subset post 2000
```{r}
Post2000Nutrients <- subset(EC_Flow.Nutrients_Wide, Date > "2000-01-01" & Date < "2020-12-31")

```

### Nutrients over Time 2000s
```{r}
Nitrogen_Time2000 <- 
  ggplot(Post2000Nutrients, aes (x=Date, y = Nitrogen))+
  geom_point()+
  geom_smooth(method=lm)
print(Nitrogen_Time2000)

 Phosphorus_Time2000 <- 
  ggplot(Post2000Nutrients, aes (x=Date, y = Phosphorus))+
  geom_point()+
  geom_smooth(method=lm)
print(Phosphorus_Time2000)
```

# Nutrient Summaries
```{r}
EC_Nutrient_MonthSummary <- EC_Flow.Nutrients_Wide %>% 
  group_by(month) %>% 
  filter(!is.na(Nitrogen), !is.na(Phosphorus)) %>% 
  summarise(meanN = mean(Nitrogen),
            sdN = sd(Nitrogen),
            minN = min(Nitrogen),
            maxN = max(Nitrogen),
            meanP = mean(Phosphorus),
            sdP = sd(Phosphorus),
            minP = min(Phosphorus),
            maxP = max(Phosphorus))

summarise(EC_Flow.Nutrients_Wide, meanN=mean(Nitrogen))

ggplot(EC_Nutrient_MonthSummary)+
  geom_point(aes(y = meanN, x=month))

ggplot(EC_Nutrient_MonthSummary)+
  geom_point(aes(y = meanP, x=month))

EC_Nutrient_YearSummary <- EC_Flow.Nutrients_Wide %>% 
  mutate(year = year(Date)) %>% 
  group_by(year) %>% 
  filter(!is.na(Mixed_N), !is.na(TP)) %>% 
  summarise(meanN = mean(Mixed_N),
            sdN = sd(Mixed_N),
            minN = min(Mixed_N),
            maxN = max(Mixed_N),
            meanP = mean(TP),
            sdP = sd(TP),
            minP = min(TP),
            maxP = max(TP))

ggplot(EC_Nutrient_YearSummary, aes(y = meanN, x=year))+
  geom_point()+
  geom_smooth(method=lm)

ggplot(EC_Nutrient_YearSummary, aes(y = meanP, x=year))+
  geom_point()+
  geom_smooth(method=lm)

```
