---
title: "Project_Code"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Set Working Directory and Load Data 
```{r setup, include=FALSE}
# Set your working directory

setwd("/Users/rmlandman/Desktop/Data Analytics/Environmental_Data_Analytics_2020/Project/EDA_2020_Project")
getwd()

# Load your packages

library(plyr)
library(tidyverse)
library(dataRetrieval)
library(lubridate)

# Set your ggplot theme
mytheme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)


# Load Ellerbe Creek Raw 

Ellerbe_Raw <- read.csv("./Data/Raw/Ellerbe_WaterQuality_Raw_Data.csv")


# Data Retrieval for USGS Flow Data 
### USGS flow data from 2 different gages (Club Blvd and Gorman)

ClubSummary <- whatNWISdata(siteNumbers = "0208675010") # What data is available? What parameters? 
EllerbeDischarge_Club <- readNWISdv(siteNumbers = "0208675010",
                           parameterCd =c("00060", "00065"), # discharge (ft3/s), gage height (ft)
                           startDate = "2008-01-01",
                           endDate = "2020-4-17")

GormanSummary  <- whatNWISdata(siteNumbers = "02086849")
EllerbeDischarge_Gorman <- readNWISdv(siteNumbers = "02086849",
                                      parameterCd = c("00060", "00065"), # discharge (ft3/s), gage height
                                      startDate = "2008-01-01",
                                      endDate = "2020-4-17")

# Save USGS Data to Raw Data Folder 
write.csv(EllerbeDischarge_Club, row.names = FALSE, 
          file = "./Data/Raw/Club_Discharge_Raw.csv")

write.csv(EllerbeDischarge_Gorman, row.names = FALSE, 
          file = "./Data/Raw/Gorman_Discharge_Raw.csv")

```

# Explore Data 
```{r}
head(Ellerbe_Raw)
colnames(Ellerbe_Raw) ### See column names to determine which to select
summary(Ellerbe_Raw$CharacteristicName) 
  ### See available characteristics to examines and number of measurments for each


```

#Clean Ellerbe Creek Data (WQ Portal)
```{r}
# Select Columns of Interest 
Ellerbe_Clean<- Ellerbe_Raw %>%
  select(OrganizationIdentifier, ActivityStartDate, ActivityStartTime.Time, MonitoringLocationIdentifier, HydrologicEvent, HydrologicCondition, CharacteristicName, ResultSampleFractionText, ResultMeasureValue,ResultMeasure.MeasureUnitCode, ResultStatusIdentifier, ResultValueTypeName) %>%
  mutate_all(na_if,"") %>% 
  drop_na(ResultMeasureValue)

write.csv(Ellerbe_Clean, row.names = FALSE, 
          file = "./Data/Processed/Ellerbe_Clean.csv")

# Explore cleaned data

head(Ellerbe_Clean)
colnames(Ellerbe_Clean)
summary(Ellerbe_Clean$ActivityTypeCode)
head(Ellerbe_Clean$ActivityTypeCode)

```

# Join USGS flow data 
```{r}

### Full Join
### Wide dataset with 2 site columns 
Ellerbe_CombinedDischarge <- full_join(EllerbeDischarge_Gorman, EllerbeDischarge_Club, by = "Date")
names(Ellerbe_CombinedDischarge)[4:7]<- c("Discharge.G","D.Approval.Code.G",
                                          "Gage.Height.G", "GH.Approval.Code.G" )
names(Ellerbe_CombinedDischarge)[10:13]<- c("Discharge.C","D.Approval.Code.C",
                                            "Gage.Height.C", "GH.Approval.Code.C")

write.csv(Ellerbe_CombinedDischarge, row.names = FALSE, 
          file = "./Data/Processed/ClubGorman_USGSDischarge_Wide.csv")


### rbind 
### Long dataset with 2 rows for each date and 1 site column 
Ellerbe_USGSDischarge_Long <- rbind(EllerbeDischarge_Club, EllerbeDischarge_Gorman)
Ellerbe_USGSDischarge_Long$Date <- as.Date(Ellerbe_USGSDischarge_Long$Date, format = "%Y-%m-%d")
names(Ellerbe_USGSDischarge_Long)[4:7]<- c("Discharge", "D.Approval.Code",
                                          "Gage.Height", "GH.Approval.Code")

write.csv(Ellerbe_USGSDischarge_Long, row.names = FALSE, 
          file = "./Data/Processed/ClubGorman_USGSDischarge_Long.csv")
```


#Pull out just nutrient data 
```{r}
Ellerbe_Nutrients <- subset(Ellerbe_Clean, CharacteristicName == "Temperature, water" | CharacteristicName == "Nitrogen, mixed forms (NH3), (NH4), organic, (NO2) and (NO3)" | CharacteristicName == "Nitrate" | CharacteristicName == "Phosphorus")


```

# Format Dates and Numeric Columns 
```{r}
Ellerbe_Nutrients$ActivityStartDate <- as.Date(Ellerbe_Nutrients$ActivityStartDate, format = "%m/%d/%y")

Ellerbe_Nutrients$ActivityStartDate <- format(Ellerbe_Nutrients$ActivityStartDate, "%y%m%d")

create.early.dates <- (function(d) {
       paste0(ifelse(d > 181217,"19", "20"),d)
       })
#
Ellerbe_Nutrients$ActivityStartDate <- create.early.dates(Ellerbe_Nutrients$ActivityStartDate)

#
Ellerbe_Nutrients$ActivityStartDate <- as.Date(Ellerbe_Nutrients$ActivityStartDate, format = "%Y%m%d")
class(Ellerbe_Nutrients$ActivityStartDate)
 

Ellerbe_Nutrients$ResultMeasureValue <-as.numeric(as.character(Ellerbe_Nutrients$ResultMeasureValue))
class(Ellerbe_Nutrients$ResultMeasureValue)

```

# Convert N and P Units 
```{r}
EC_Nutrients_Units <- Ellerbe_Nutrients %>%
  group_by (MonitoringLocationIdentifier, ActivityStartDate, HydrologicEvent, CharacteristicName,ResultSampleFractionText, ResultMeasure.MeasureUnitCode) %>% 
  summarise (meanresult = mean(ResultMeasureValue))  %>% 
  spread(ResultMeasure.MeasureUnitCode, meanresult) 

# Convert NO3 to N
### EC_Nutrients_Units <- mutate(EC_Nutrients_Units, Nitrogen = `mg/l asNO3`*0.2258)

#Convert PO4 to P 
### EC_Nutrients_Units <- mutate(EC_Nutrients_Units, Phosphorus = `mg/l PO4`*0.3261)

# Just need to remove the columns for NO3 and PO4 because dataset already has columns for just N and P for each sample so no need to convert 
EC_Nutrients_Units <- subset(EC_Nutrients_Units, select = -c(7,11:13))

colnames(EC_Nutrients_Units)

# Gather data 
####Ellerbe_Nutrients_Processed <- gather(EC_Nutrients_Units, "Unit", "Measurment", c( `mg/l`, `mg/l as P`, `deg C`, `mg/l as N`))


###Ellerbe_Nutrients_Processed <- unite(EC_Nutrients_Units, col = "ResultMeasureValue", `mg/l`, `mg/l as P`, `deg C`, `mg/l as N`, rm.na=TRUE)

Ellerbe_Nutrients_Processed <- pivot_longer(EC_Nutrients_Units, cols = c(`mg/l`, `mg/l as P`, `deg C`, `mg/l as N`), names_to = c("Units")) %>% 
  drop_na(value)

```


#Save Processed Long Data
```{r}

write.csv(Ellerbe_Nutrients_Processed, row.names = FALSE, 
          file = "./Data/Processed/Ellerbe_Nutrients_Processed.csv")
```

#Convert to Wide - All Locations

```{r}
Ellerbe_Nutrients_Wide <- Ellerbe_Nutrients_Processed %>%
  group_by (MonitoringLocationIdentifier, ActivityStartDate, HydrologicEvent, CharacteristicName,ResultSampleFractionText) %>% 
  spread(CharacteristicName, value) %>% 
  mutate(month = month(ActivityStartDate, label = TRUE)) %>% 
  separate(MonitoringLocationIdentifier, c("agency", "site_no"), sep = "-") %>% 
  rename("Date" = ActivityStartDate, "SampleFraction" = ResultSampleFractionText, "Nitrate" = Nitrate, "Nitrogen" = `Nitrogen, mixed forms (NH3), (NH4), organic, (NO2) and (NO3)`,"Temperature" =`Temperature, water`)


# Save processed dataset 
write.csv(Ellerbe_Nutrients_Wide, row.names = FALSE, 
          file = "./Data/Processed/Ellerbe_Nutrients_Wide_Processed.csv")

```

# Subset Wide Nutrient Data to include just measurements at Gages 

```{r}

ClubGorman_Nutrients_Wide <- subset(Ellerbe_Nutrients_Wide, site_no == "0208675010" | site_no == "02086849" | site_no == "J1330000" | site_no == "J1270000"| site_no == "0208675009")

write.csv(ClubGorman_Nutrients_Wide, row.names = FALSE, 
          file = "./Data/Processed/ClubGorman_Nutrients_Wide_Processed.csv")

```

# Subset Long Nutrient Data to include just measurements at Gages 
```{r}
Ellerbe_Nutrients_Long <- Ellerbe_Nutrients_Processed %>%
  group_by (MonitoringLocationIdentifier, ActivityStartDate, HydrologicEvent, CharacteristicName,ResultSampleFractionText) %>% 
  mutate(month = month(ActivityStartDate, label = TRUE)) %>% 
  separate(MonitoringLocationIdentifier, c("agency", "site_no"), sep = "-") %>% 
  rename("Date" = ActivityStartDate, "SampleFraction" = ResultSampleFractionText)

ClubGorman_Nutrients_Long <- subset(Ellerbe_Nutrients_Long,  site_no == "0208675010" | site_no == "02086849" | site_no == "J1330000" | site_no == "J1270000"| site_no == "0208675009")

write.csv(ClubGorman_Nutrients_Long, row.names = FALSE, 
          file = "./Data/Processed/ClubGorman_Nutrients_Long_Processed.csv")
```


# Join Discharge and Nutrients 
```{r}

### Just USGS Gage Sites - Wide 
class(ClubGorman_Nutrients_Wide$Date)
class(Ellerbe_USGSDischarge_Long$Date)

EC_USGS_Combined <- full_join(ClubGorman_Nutrients_Wide, Ellerbe_USGSDischarge_Long, by = c("site_no","Date"))

EC_USGS_Combined$Location <- EC_USGS_Combined$site_no

EC_USGS_Combined$Location <- gsub("J1330000", "Downstream", EC_USGS_Combined$Location)
EC_USGS_Combined$Location <- gsub("02086849", "Downstream", EC_USGS_Combined$Location)
EC_USGS_Combined$Location <- gsub("0208675010", "Upstream", EC_USGS_Combined$Location)
EC_USGS_Combined$Location <- gsub("J1270000", "Upstream", EC_USGS_Combined$Location)
EC_USGS_Combined$Location <- gsub("0208675009", "Upstream", EC_USGS_Combined$Location)


write.csv(EC_USGS_Combined, row.names = FALSE, 
          file = "./Data/Processed/ClubGorman_Flow.Nutrients_Wide_Processed.csv")

# Just USGS Gage Sites - Wide 
class(ClubGorman_Nutrients_Long$Date)
EC_USGS_Long_Combined <- full_join(ClubGorman_Nutrients_Long, Ellerbe_USGSDischarge_Long, by = c("site_no","Date"))

EC_USGS_Long_Combined$Location <- EC_USGS_Long_Combined$site_no

EC_USGS_Long_Combined$Location <- gsub("J1330000", "Downstream", EC_USGS_Long_Combined$Location)
EC_USGS_Long_Combined$Location <- gsub("02086849", "Downstream", EC_USGS_Long_Combined$Location)
EC_USGS_Long_Combined$Location <- gsub("0208675010", "Upstream", EC_USGS_Long_Combined$Location)
EC_USGS_Long_Combined$Location <- gsub("0208675010", "Upstream", EC_USGS_Long_Combined$Location)
EC_USGS_Long_Combined$Location <- gsub("J1270000", "Upstream", EC_USGS_Long_Combined$Location)
EC_USGS_Long_Combined$Location <- gsub("0208675009", "Upstream", EC_USGS_Long_Combined$Location)

EC_USGS_Long_Combined$CharacteristicName <- gsub("`Nitrogen, mixed forms (NH3), (NH4), organic, (NO2) and (NO3)`", "Nitrogen", EC_USGS_Long_Combined$CharacteristicName)

write.csv(EC_USGS_Long_Combined, row.names = FALSE, 
          file = "./Data/Processed/ClubGorman_Flow.Nutrients_Long_Processed.csv")

### Join Flow data with all Nutrient data 

EC_Flow.Nutrients_Wide <- full_join(Ellerbe_Nutrients_Wide, Ellerbe_CombinedDischarge, by = "Date")

write.csv(EC_Flow.Nutrients_Wide, row.names = FALSE, 
          file = "./Data/Processed/EC_Flow.Nutrients_Wide_Processed.csv")


```



